### 2023-04-28 paper.tex
Make some proposed changes, add to intro, write conclusion

\section{Introduction}
This paper will take a look at three popular language models utilizing
transformers in their architecture. It will also give a quick overview
of some applications and additions to the generic model. The aim is to
give the reader a general overview of the current landscape of different
transformer based language model architectures and some example applications.

The three different models are all similar in architecture being mostly
faithful to the original transformer as presented in 
\cite{vaswani_attention_2017}. However, each of them have a different approach
to using the decoder and encoder stacks of the transformer. GPT uses only
the decoder, BERT only the encoder and T5 both the encoder and decoder
\cite{radford_improving_nodate,devlin_bert_2019,raffel_exploring_2020}.

@@

The objective of the unsupervised pre-training stage is to teach the model
general language understanding. The exact details between models varies but
in general, a massive number tokens are fed into to model during the course
of pre-training. Each word corresponds to a single or a few tokens depending on 
the encoding used \cite{noauthor_pricing_nodate} and the number of tokens
used in training can go up to the trillions \cite{raffel_exploring_2020}. 
The model is asked to predict a missing token based on context.

@@

\subsection{Attention masking}
Attention masking prevents the model from attending to some tokens in the input. At any
given time $j$, the model is allowed to see input up to index $i$. Masking is used to prevent the model from
attending to future input which would trivialize training. \cite{raffel_exploring_2020}

The original transformer's encoder-decoder architecture as specified in 
\cite{vaswani_attention_2017} corresponds to the leftmost masking pattern as
specified in figure \ref{masking_patterns}. T5 uses this architecture and masking pattern
\cite{raffel_exploring_2020}. As previously mentioned, BERT uses an encoder
only architecture \cite{devlin_bert_2019} and uses a fully-visible masking 
pattern with a special classification token \cite{raffel_exploring_2020}. GPT on
the other hand, uses a decoder only architecture \cite{radford_improving_nodate}
and uses causal masking \cite{liu_generating_2018}.

@@

Calculating the entire attention history for a long sequence can be an expensive operation.
It is possible to use a sparse version of the attention masking pattern and still achieve 
comparable performance to a full attention mask. A sparse transformer with a fixed attention
pattern includes cells that can summarize the previous locations and propagate 
that forward. This allows computing longer sequences of input tokens and is used in GPT-3
\cite{brown_language_2020}. An illustrating example of different masking patterns
can be seen in figure \ref{sparse_transformer}. \cite{child_generating_2019}

@@

As an example, \cite{brown_language_2020} used English to French translation.
The researchers gave the model between 10 to 100 pairs of examples of the
same sentence in English and French. The last pair had just an English sentence
and the model was expected to produce the corresponding sentence translated into
French.

While few-shot learning requires less data and computing power, which makes it more
accessible compared to fine-tuning a large model, at least during the writing of 
\cite{brown_language_2020} failed to produce comparable results to fine-tuning.
The number of examples given is restricted by the size of the model's context,
which for GPT-3 was 2048 tokens, around 1500 words \cite{noauthor_pricing_nodate}.

\subsection{Bidirectional Encoder Representations from Transformers}
Bidirectional Encoder Representations from Transformers, henceforth referred to as BERT,
is a language model utilizing masking of words during training to achieve bidirectionality
in the model. \cite{devlin_bert_2019}

BERT uses a similar architecture as the original transformer. It uses the encoder with
multiple layers. The main difference is in the training methods. In the pre-training step,
two different training approaches are used: masked language model and next sentence prediction.
Masking allows the self-attention layer to take into account both sides of a given word, instead
of just one side, as in GPT which can only use the previous words as context. Using next sentence
prediction improves the model's understanding of the relationships between sentences. This is
useful for question answering and natural language inference type of tasks. \cite{devlin_bert_2019}

A masked language model (MLM) is achived by hiding a certain number of tokens, 12\% in the paper, and
asking the model to predict the hidden token. The model can then use both sides of the token as
context. Some of the tokens marked for masking are instead turned 
into random tokens or kept as is. This helps prevent a difference between pre-training and fine-tuning.
Other unidirectional methods predict the entire text during training while MLM only 
does so only for the masked tokens. \cite{devlin_bert_2019}

@@

The main differentiating factor of T5 is the way it chooses to model tasks. Instead of 
tailoring each output to the specific task at hand, every task takes in a text prefix
describing the task and outputting the results as text. For example, for STS-B classification
task the model is trained to output one of a set of floating point numbers 
$x \in \{ 1+0.2k \mid k \in \{ 0, 1, ..., 20\} \}$ as a text. Many other models just 
output a number directly as a sliding scale from 1 to 5.

\section{Applications}

@@

\section{Problems}
\subsection{Biases}
The writers of \cite{brown_language_2020} did a review of gender, racial and religious
biases in the model they presented in the paper. Their conclusion was that there were
certain biases in place which matched the biases present in the training data, that being
the internet.

@@

>>>>

\section{Introduction}

TODO: write a generic paragraph

Transformers can be split into two main parts: an encoder and a decoder stack. 
The encoder reads the input text converting it into a vector representation. This is
then fed into the decoder which then generates the output utilizing this vector.
It is not necessary to have both an encoder and a decoder in the transformers
and the same tasks can be performed even without one of these present.

While encoder-decoder language models existed before the transformer, it was 
the first one to use only an attention mechanism for all computation. 
The attention mechanism is a mapping for a query and key-value pairs to some output. 
Intuitively, in transformers
it is used by the model to give different weights for different words in the input when
determining what the output should be [TODO: citation needed].

This paper will take a look at three popular language models utilizing
transformers in their architecture. It will also give a quick overview
of some applications and additions to the generic model. The aim is to
give the reader a general overview of the current landscape of different
transformer based language model architectures and some example applications.

The three different models studied in this paper are all similar in architecture 
being mostly faithful to the original transformer as presented in 
\cite{vaswani_attention_2017}. However, each of them have a different approach
to using the decoder and encoder stacks of the transformer. GPT uses only
the decoder, BERT only the encoder and T5 both the encoder and decoder
\cite{radford_improving_nodate,devlin_bert_2019,raffel_exploring_2020}.

@@

The objective of the unsupervised pre-training stage is to teach the model
general language understanding. The exact details between models varies but
in general, a massive number of tokens are fed into the model during the course
of pre-training. Each word corresponds to a single or a few tokens depending on 
the encoding used \cite{noauthor_pricing_nodate} and the number of tokens
used in training can go up to the trillions \cite{raffel_exploring_2020}. 
The model is asked to predict a missing token based on context.

@@

\subsection{Attention masking}
Attention masking prevents the model from seeing some tokens in the input. At any
given time $j$, the model is allowed to use input up to index $i$. Masking is used to prevent the model from
seeing future input which would trivialize training. \cite{raffel_exploring_2020}

The original transformer's encoder-decoder architecture as specified in 
\cite{vaswani_attention_2017} corresponds to the leftmost masking pattern as
specified in Figure \ref{masking_patterns}. T5 uses this architecture and masking pattern
\cite{raffel_exploring_2020}. As previously mentioned, BERT uses an encoder
only architecture \cite{devlin_bert_2019} and uses a fully-visible masking 
pattern with a special classification token \cite{raffel_exploring_2020}. GPT on
the other hand, uses a decoder only architecture \cite{radford_improving_nodate}
and uses causal masking \cite{liu_generating_2018}.

@@

Calculating the entire attention history for a long sequence can be an expensive operation.
It is possible to use a sparse version of the attention masking pattern and still achieve 
comparable performance to a full attention mask. A sparse transformer with a fixed attention
pattern includes cells that can summarize the previous locations and propagate 
that forward. This allows computing longer sequences of input tokens and is used in GPT-3
\cite{brown_language_2020}. An illustrating example of different masking patterns
can be seen in Figure \ref{sparse_transformer}. \cite{child_generating_2019}

@@

As an example, Brown et al. \cite{brown_language_2020} used English to French translation.
The researchers gave the model between 10 to 100 pairs of examples of the
same sentence in English and French. The last pair had just an English sentence
and the model was expected to produce the corresponding sentence translated into
French.

While few-shot learning requires less data and computing power, which makes it more
accessible compared to fine-tuning a large model, Brown et al. \cite{brown_language_2020} 
were unable to produce comparable results to fine-tuning.
The number of examples given is restricted by the size of the model's context,
which for GPT-3 was 2048 tokens, around 1500 words \cite{noauthor_pricing_nodate}.

\subsection{Bidirectional Encoder Representations from Transformers}
Bidirectional Encoder Representations from Transformers (BERT)
is a language model utilizing masking of words during training to achieve bidirectionality
in the model. \cite{devlin_bert_2019}

BERT uses a similar architecture as the original transformer. It uses the encoder with
multiple layers. The main difference is in the training methods. In the pre-training step,
two different training approaches are used: masked language model and next sentence prediction.
Word masking allows the self-attention layer to take into account both sides of a given word, instead
of just one side, as in GPT which can only use the previous words as context. Word masking is
distinct from attention masking and refers to the hiding of individual words in the training data
before feeding it into the model.
% TODO: verify this
 Using next sentence
prediction improves the model's understanding of the relationships between sentences. This is
useful for question answering and natural language inference type of tasks. \cite{devlin_bert_2019}

A masked language model (MLM) is achived by hiding a certain number of tokens and
asking the model to predict the hidden token. The model can then use both sides of the token as
context. Some of the tokens marked for masking are instead turned 
into random tokens or kept as is. This helps prevent a difference between pre-training and fine-tuning.
Other unidirectional methods predict the entire text during training while MLM only 
does so only for the masked tokens. \cite{devlin_bert_2019}

@@

The main differentiating factor of T5 is the way it chooses to model tasks. Instead of 
tailoring each output to the specific task at hand, every task takes in a text prefix
describing the task and outputting the results as text. For example, for a classification
task the model is trained to output one of a set of floating point numbers 
$x \in \{ 1+0.2k \mid k \in \{ 0, 1, ..., 20\} \}$ as text. Many other models just 
output a number directly as a sliding scale from 1 to 5.

\section{Extensions}

@@

\section{Problems}
\subsection{Biases}
Brown et al. \cite{brown_language_2020} did a review of gender, racial and religious
biases in the model they presented in the paper. Their conclusion was that there were
certain biases in place which matched the biases present in the training data, that being
the internet.

@@

\section{Conclusion}
This paper introduced three popular language models based on the transformer architecture.
They are mostly similar, all faithful to the original transformer architecture, with
some key differences in the usage of attention masking and encoder-decoder stacks.

Transformers are able to produce impressive results on many different domains of natural
language processing. However, it is important to keep in mind that they are not without 
their problems. As with any machine learning model,
biases should be considered and the outputs validated.

Hopefully this paper serves as a good introduction to the general landscape of 
transformers in language models and helps in understanding what is going on in
large language models.
